The Form of Uncertainty Affects Selection for Social Learning

### 2023-02-23 ms.tex
revisions nearly if not totally done

\begin{abstract} 
Social learning is a critical adaptation for dealing with the variable world. However, variability
takes different forms. Uncertainty is a severe form of variability where the space of possible
decisions or probabilities of associated outcomes are unknown.
%"probabilities of decisions" didn't make sense to me so I rephrased, but hopefully still accurate.
We identified four distinct and theoretically-important sources of uncertainty: temporal
environmental variability, payoff ambiguity, selection-set size, and effective lifespan. All these
kinds of uncertainty make it nearly impossible to fully learn about variability. We develop an
evolutionary agent-based model to test how each form of uncertainty affects the evolution of social
learning. Agents perform one of several behaviors, modeled as a multi-armed bandit, to acquire
payoffs. \edit{Agents explore and learn about behavioral payoffs individually through an adaptive
cognitive model that uses a softmax decision rule for behavior choice.} Use of
vertical and oblique \edit{payoff-biased} social learning \edit{evolved} to serve as a scaffold \edit{for
adaptive individual learning, meaning social and individual learning complement one
another---they are not opposite strategies.} We found that different types of uncertainty had varying effects.
Temporal environmental variability suppressed social learning, whereas larger selection-set size
promoted social learning, even when the environment changed frequently. Payoff ambiguity and lifespan
had more complex effects on social learning, dependent on interactions with other uncertainty
parameters. \edit{By assuming adaptive individual-level cognition, this study begins to explain how
social learning can predominate despite high levels of various forms of uncertainty in the
real-world.}

  % the importance of clearly operationalizing uncertainty for understanding its role in cultural evolution.
% Social learning is essential to human survival. It is likely to evolve when it is more
% efficient than asocial, trial-and-error learning. Theoretically, social learning
% is adaptive for some uncertainty, but too much uncertainty makes
% social information unreliable. This fact is empirically
% supported across biology and human sciences. However, it is unclear how
% specific classes and types of uncertainty affect social
% learning evolution. Furthermore, existing models and experimental operationalizations of
% uncertainty are ambiguously related, and 
% tend to only consider a small number of behavioral
% choices.  Here we use evolutionary agent-based modeling to consolidate
% models of uncertainty in social learning evolution to address these issues.
% We model societies of agents who perform one of a number
% of possible behaviors to acquire payoffs in a time-varying environment.
% We identified complex patterns in social learning evolution depending on contextual
% uncertainty variables: environmental variability, ambiguous payoffs, 
% decision sets size, and effective lifespans. 
% Our work advances social learning evolution theory in a way that could help 
% guide human, non-human, and artificial intelligence towards optimal responses 
% to existential threats and new opportunities.
% \footnote{This document contains
% \total{citnum} references.}  
\end{abstract}

@@

\begin{figure}
  \caption{Agents are randomly initialized as social learners or not, with their
  payoff observations all initialized to zero (A). Then agents begin selecting
and performing behaviors and accumulating payoffs, which goes on for $L$
timesteps (B). After $L$ time steps, agents are selected to reproduce,
social learner children learn from a member of their parent's generation, and
the previous generation dies off (C). The simulation stops if children are all
social or asocial learners (i.e.\ the system reaches fixation).
Otherwise it repeats another generation and evolution continues.} 
% \cm{What does (3) denote at end of Fig 1 caption? Does having only green agents in Fig1B give the impression that only social learners are learning within lifetime? Could the agent be half green/purple? Is it weird that in Fig1C a teacher with relatively low payoffs is being selected? Could be fixed by having the dotted line go from rich purple women's head to green baby's. }
  \label{fig:schematic}
  \centering
    \includegraphics[width=\textwidth]{Figures/IntraInterGenerationalDynamics.pdf}
\end{figure}

@@

        $\meansl$ & Mean social-learning prevalence over agents (Fig.~\ref{fig:mainResults})
                  & $ [0.0, 1.0]$ \\

        $\meanG$ & Mean number of generations to convergence (Fig.~\ref{fig:steps}) & \{1,2,\ldots\} \\

@@

Our main results are illustrated by Figure~\ref{fig:mainResults}, which shows the
proportion of simulation runs that fixated at 100\% social learners as a function of
each of our four uncertainty measures. First, we found that reliance on social
learning monotonically decreases as temporal environmental variability, $u$,
increases. When the environment is more likely to change between generations,
information learned from the previous generation is less likely to be of value,
decreasing selection for social learning. \edit{Past some threshold, social learning never evolves.  We define this threshold environmental variability as the \emph{social learning extinction variability} (Figure~\ref{fig:extinctionVariability}).} 
However, the exact nature of the relationship between social
learning and temporal environmental variability was moderated by the other three uncertainty parameters.

@@

\begin{figure}
  \caption{As the number of
  behaviors afforded by the environment increases (x-axis), 
  social learning is supported up to increasingly large values of environmental variability (SL extinction
    varibility, y-axis).}
  \label{fig:extinctionVariability}
  \centering
    \includegraphics[width=0.8\textwidth]{Figures/SL_Extinction_Variability.pdf}
\end{figure}

>>>>

\begin{abstract} 
Social learning is a critical adaptation for dealing with different forms of variability. 
Uncertainty is a severe form of variability where the space of possible
decisions or probabilities of associated outcomes are unknown.
We identified four theoretically-important sources of uncertainty: temporal
environmental variability, payoff ambiguity, selection-set size, and effective lifespan. When these combine,
it is nearly impossible to fully learn about the environment. We develop an
evolutionary agent-based model to test how each form of uncertainty affects the evolution of social
learning. Agents perform one of several behaviors, modeled as a multi-armed bandit, to acquire
payoffs. \edit{All agents learn about behavioral payoffs individually through an adaptive
behavior-choice model that uses a softmax decision rule.} Use of
vertical and oblique \edit{payoff-biased} social learning \edit{evolved} to serve as a scaffold \edit{for
adaptive individual learning---they are not opposite strategies}. Different types of uncertainty had varying effects.
Temporal environmental variability suppressed social learning, whereas larger selection-set size
promoted social learning, even when the environment changed frequently. Payoff ambiguity and lifespan
interacted with other uncertainty parameters. \edit{This study begins to explain how
social learning can predominate despite highly variable real-world environments when effective individual learning helps individuals recover from learning outdated social information.}
%, meaning social and individual learning complement one another
  % the importance of clearly operationalizing uncertainty for understanding its role in cultural evolution.
% Social learning is essential to human survival. It is likely to evolve when it is more
% efficient than asocial, trial-and-error learning. Theoretically, social learning
% is adaptive for some uncertainty, but too much uncertainty makes
% social information unreliable. This fact is empirically
% supported across biology and human sciences. However, it is unclear how
% specific classes and types of uncertainty affect social
% learning evolution. Furthermore, existing models and experimental operationalizations of
% uncertainty are ambiguously related, and 
% tend to only consider a small number of behavioral
% choices.  Here we use evolutionary agent-based modeling to consolidate
% models of uncertainty in social learning evolution to address these issues.
% We model societies of agents who perform one of a number
% of possible behaviors to acquire payoffs in a time-varying environment.
% We identified complex patterns in social learning evolution depending on contextual
% uncertainty variables: environmental variability, ambiguous payoffs, 
% decision sets size, and effective lifespans. 
% Our work advances social learning evolution theory in a way that could help 
% guide human, non-human, and artificial intelligence towards optimal responses 
% to existential threats and new opportunities.
% \footnote{This document contains
% \total{citnum} references.}  
\end{abstract}

@@

\begin{figure}
  \caption{Agents are randomly initialized as social learners or not, with their
  payoff observations all initialized to zero (A). Then agents begin selecting
and performing behaviors and accumulating payoffs, which goes on for $L$
timesteps (B). After $L$ time steps, agents are selected to reproduce,
social learner children learn from a member of their parent's generation, and
the previous generation dies off (C). The simulation stops if children are all
social or asocial learners (i.e.\ the system reaches fixation).
Otherwise it repeats another generation and evolution continues.} 
% \cm{What does (3) denote at end of Fig 1 caption? Does having only green agents in Fig1B give the impression that only social learners are learning within lifetime? Could the agent be half green/purple? Is it weird that in Fig1C a teacher with relatively low payoffs is being selected? Could be fixed by having the dotted line go from rich purple women's head to green baby's. }
  \label{fig:schematic}
  \centering
    \includegraphics[width=1.05\textwidth]{Figures/IntraInterGenerationalDynamics.pdf}
\end{figure}

@@

        $\meansl$ & Social learning fixation frequency (Fig.~\ref{fig:mainResults})
                  &  $[0.0, 1.0]$ \\

        $\meanG$ & Mean time to convergence in units of generations (Fig.~\ref{fig:steps}) & \{1,2,\ldots\} \\

@@

Our main results are illustrated by Figure~\ref{fig:mainResults}, which shows the
proportion of simulation runs that fixated at 100\% social learners as a function of
each of our four uncertainty measures. First, we found that reliance on social
learning monotonically decreases as temporal environmental variability, $u$,
increases \edit{(Figure~\ref{fig:mainResults})}. 
When the environment is more likely to change between generations,
information learned from the previous generation is less likely to be of value,
decreasing selection for social learning.  
However, the exact nature of the relationship between social
learning and temporal environmental variability was moderated by the other three uncertainty parameters.

\edit{Across all parameter settings, we observed that there exists some threshold value of environmental
  variability beyond which the evolution of social learning is exceedingly rare.  We call this the
  \emph{social learning extinction variability}, denoted $u_\text{e}$ (Figure~\ref{fig:extinctionVariability}), which is expressed formally as 
\begin{equation}
  u_\text{e} = \min~\{ u~|~\meansl < 0.01 \}.
  \label{eq:extinctionVariability}
\end{equation}
\noindent
In other words, $u_\text{e}$ is the smallest value of $u$ for which the social learning fixated
less than one in one hundred times. We will use this auxiliary measure to analyze the effect of
the other uncertainty parameters and to highlight that social learning frequently 
evolves despite high levels of environmental variability.}

@@

\begin{figure}
  \caption{As the selection-set size, $B$, and effective lifespan, $L$, increase (set equal along x-axis), 
  social learning frequently evolves across an increasingly large range of
  environmental variability values, i.e., up to
  increasingly large values of social learning extinction variability
(y-axis; defined in Equation~\ref{eq:extinctionVariability}).}
  \label{fig:extinctionVariability}
  \centering
    \includegraphics[width=0.8\textwidth]{Figures/SL_Extinction_Variability.pdf}
\end{figure}
